---
title: "DBSCAN and OPTICS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DBSCAN and OPTICS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load package

```{r setup}
library(haufenR)
```

## Overview

This vignette shows how to use the density-based clustering methods `dbscan()` and `optics()` from the `haufenR` package. The focus is on usage and interpretation of the output. We use a small two-dimensional example data set with two compact groups and two additional points outside these groups.

For a full description of the returned values, see `?dbscan` and `?optics`.

```{r example-data}
x <- rbind(
  c(0.00, 0.00),
  c(0.05, 0.02),
  c(0.12, 0.09),
  c(0.20, 0.01),
  c(0.31, 0.14),
  c(0.40, 0.05),

  c(1.55, 1.00),
  c(1.68, 1.11),
  c(1.83, 1.05),
  c(1.97, 1.20),
  c(2.12, 1.12),
  c(2.28, 1.30),

  c(1.00, 0.55),
  c(2.95, 0.15)
)

plot(x, pch = 19, main = "Example data")
```

## DBSCAN

`dbscan()` groups together points in dense regions and labels isolated points as noise.

The main parameters are `eps` and `minPts`:

- `eps` is the neighborhood radius.  
- `minPts` is the minimum number of points needed for a core point.

```{r dbscan-example}
res_db <- dbscan(x, eps = 0.25, minPts = 3)

print(res_db)
```

The result can be visualized with the S3 plot method.

```{r dbscan-plot}
plot(res_db, x)
```

In this example, the points on the left form one dense group, the points in the upper-right region form another group, and the two remaining points are treated as noise.

In the plot: filled points are core points, open circles are border points, crosses indicate noise.

## Inspecting the DBSCAN result

The compact `print()` output is useful for a quick overview. The returned object also contains extra data that can be accessed directly.

```{r dbscan-fields}
table(res_db$clusters)

res_db$core

res_db$clusters_count

res_db$noise_count

res_db$cluster_sizes

res_db$core_count

res_db$border_count
```

- `clusters`: integer vector of length `n`, where `0` means noise.
- `core`: logical vector of length `n`, where `TRUE` means a core point.
- `clusters_count`, `noise_count`: short summary for quick inspection.
- `cluster_sizes`: cluster sizes without noise. 
- `core_count`, `border_count`: number of core and border points.

## OPTICS

`optics()` does not directly return cluster labels. Instead, it returns an ordering of the points together with reachability distances (density).

The main parameters are also `eps` and `minPts`:

- `eps` is the neighborhood radius.  
- `minPts` is the minimum number of points needed for a core point.

```{r optics-example}
res_opt <- optics(x, eps = 0.30, minPts = 3)

print(res_opt)
```

A common way to inspect the result is the reachability plot.

```{r optics-plot}
plot(res_opt)
```

Low bars in the reachability plot indicate dense regions. Higher bars or `Inf` values indicate transitions between regions or start points of new regions.

## Inspecting the OPTICS result

As with DBSCAN, the compact `print()` output is only a summary. The returned object also stores the raw ordering and distance information.

```{r optics-fields}
res_opt$order

res_opt$reachability[res_opt$order]

res_opt$core_dist

res_opt$core_count

res_opt$start_points_count
```

- `order`: the processing order returned by OPTICS.
- `reachability`: reachability distance for each point.
- `core_dist`: core distance for each point.
- `core_count`, `start_points_count`: short summary.

## Remarks

Both functions currently compute a full distance matrix, so they are **O(n^2)** in time and memory and may become slow for large data sets.

For large inputs, they may also issue a warning before performing this computation.
