---
title: "Haufenbildung"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Haufenbildung}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Package "HaufenR"

**Introduction** <br/>
Cluster analysis (German: "Haufenbildung") is a method to discover similarity structures in data. The goal is to group data points so that points within a group are similar to each other and distinct from points in other clusters.


The Package "**haufenR**" implements several clustering methods:<br/>
1. `k_means`: Partitions data into K clusters based on mean values <br/>
2. `k_medoids`: A robust alternative to k-Means, using medoids as cluster centers<br/>
3. `spectral_clustering`: Uses eigenvalues of the similarity matrix to identify clusters<br/>
4. `hc_hierarchical_clustering`: Produces a dendritic structure of clusters<br/>
5. `dbscan`/`optics`: Density-based clustering methods for detecting clusters of arbitrary shape<br/>

```{r setup}
library(haufenR)
```

---

## 1. k-Means Clustering
`k-means` partitions numeric data into K clusters by iteratively assigning points to the nearest center and updating the centers until they stabilize. Our `k_means` - function allows multiple random starts to help finding a better solution.

### Example 1: Random Clusters
We generate a random data set
```{r}
set.seed(123)

X <- rbind(matrix(rnorm(50*2, mean = 0), ncol = 2),
           matrix(rnorm(50*2, mean = 3), ncol = 2),
           matrix(rnorm(50*2, mean = 6), ncol = 2))

head(X)
```
Then we apply `k_means` from "haufenR"
```{r}
res <- k_means(X, K = 3, max_iter = 100, nstart = 5)

res$centers

table(res$clusters)
```
Visualize
```{r}
plot(X, col = res$clusters, pch = 19, cex = 0.2, main = "k-means Cluster")
points(res$centers, col = 1:3, pch = 8, cex = 2) # Mark centers
```

### Example 2: Comparison with R's `kmeans`

We use `kmeans` and `k_means` on the same data set and compare
```{r}
set.seed(123)

res_r <- kmeans(X, centers=3) #apply kmeans from R
```

Compare plots
```{r}
par(mfrow = c(1,2)) #Side-by-side plotting

# Visualize own k_mean
plot(X, col = res$clusters, pch = 19, cex = 0.2, main = "Own k_means")
points(res$centers, col = 1:3, pch = 8, cex = 2)

# Visualize R's built in kmean
plot(X, col = res_r$cluster, pch = 19, cex = 0.2, main = "R kmeans")
points(res_r$centers, col = 1:3, pch = 8, cex = 2)
```

Compare centers
```{r}
# Own centers
res$centers
# R's centers
res_r$centers #Same centers (with same seed)

par(mfrow = c(1,1)) #Reset plotting
```

Result: Our own written `k_means` does the same as `kmeans`

### Example 3: Another dataset
```{r}
set.seed(42)
X2 <- rbind(matrix(rnorm(302, mean = 0), ncol = 2),
           matrix(rnorm(502, mean = 5), ncol = 2))
res2 <- k_means(X2, K = 2)
```

Visualize
```{r}
plot(X2, col = res2$clusters, pch=19, cex = 0.2)
points(res2$centers, col = 1:2, pch = 8, cex = 2)
```

### Example 4: Edge case with a single point
```{r}
single_point <- matrix(c(1,2), nrow = 1)
res_single <- k_means(single_point, K = 1)

res_single$centers
res_single$clusters

plot(single_point, col = res_single$clusters, pch = 19, main = "k-means Cluster")
points(res_single$centers, col = 1, pch = 8, cex = 2)
```

### Example 5: Demonstrating multiple random starts improves within-cluster of sum squares
We use `k_means` on a data set with 1 start and with 5 starts
```{r}
set.seed(123)
X3 <- rbind(matrix(rnorm(50*2, mean = 0, sd = 1.5), ncol = 2),
           matrix(rnorm(50*2, mean = 2, sd = 1.5), ncol = 2),
           matrix(rnorm(50*2, mean = 4, sd = 1.5), ncol = 2))
res1 <- k_means(X3, K = 3, nstart = 1)
res5 <- k_means(X3, K = 3, nstart = 5)
```
Compare the tot_withinss
```{r}
res1$tot_withinss
res5$tot_withinss
```
Result: the total amount of distances of the data points to the centers inside each cluster is smaller with a bigger amount of starts, which proves we have have found a way to potentially optimize our code

### Example 6: Invalid input
```{r}
invalid_X <- matrix(c("a", "b", "c", "d"), ncol = 2)
tryCatch(
  k_means(invalid_X, K = 2),
  error = function(e) message("Caught expected error: ", e$message)
)
```
Result: Our Code is able to throw an error if the input is wrong

<br/>

```{r}
library(haufenR)
library(cluster)
```

## 2. K-Medoids Clustering

The k-medoids algorithm is a clustering method similar to k-means,
but instead of using the mean of a cluster it chooses actual data points
(medoids) as cluster centers.

This makes the algorithm more robust to outliers compared to k-means.

This vignette demonstrates how to use the k_medoids() function from the
haufenR package and compares it with the pam() implementation from
the cluster package.

### Example 1: Random clusters

We first generate an artificial dataset consisting of three clusters.

```{r}
set.seed(123)

X <- rbind(
  matrix(rnorm(50*2, mean = 0), ncol = 2),
  matrix(rnorm(50*2, mean = 3), ncol = 2),
  matrix(rnorm(50*2, mean = 6), ncol = 2)
)

head(X)
```

Apply our own implementation of the k-medoids algorithm.

```{r}
res <- k_medoids(X, k = 3)

res$medoids
table(res$clusters)

```

### Visualization
```{r}
plot(X, col = res$clusters, pch = 19,
     main = "Own k_medoids implementation", cex.main = 0.8)

points(X[res$medoids, ], pch = 8, cex = 2)

```

The star symbols indicate the selected medoids.

### Example 2: Comparison with PAM

We now compare our implementation with the pam() algorithm from the
cluster package.

```{r}

pam_res <- pam(X, k = 3)

par(mfrow = c(1,2))

plot(X, col = res$clusters, pch = 19,
     main = "Own k_medoids", cex.main = 0.8)
points(X[res$medoids, ], pch = 8, cex = 2)

plot(X, col = pam_res$clustering, pch = 19,
     main = "cluster::pam", cex.main = 0.8)
points(X[pam_res$medoids, ], pch = 8, cex = 2)

par(mfrow = c(1,1))

```

Although the exact cluster assignments may differ due to random
initialization, both algorithms typically produce very similar
cluster structures.

### Example 3: Another dataset

```{r}

set.seed(42)

X2 <- rbind(
  matrix(rnorm(50*2, mean = 0), ncol = 2),
  matrix(rnorm(50*2, mean = 5), ncol = 2)
)

res2 <- k_medoids(X2, k = 2)

plot(X2, col = res2$clusters, pch = 19,
     main = "Two clusters")

points(X2[res2$medoids, ], pch = 8, cex = 2)

```

### Example 4: Single-point dataset

The algorithm also works with minimal datasets.

```{r}

single_point <- matrix(c(1,2), nrow = 1)

res_single <- k_medoids(single_point, k = 1)

res_single$medoids
res_single$clusters

```

### Example 5: Invalid input

Our implementation checks whether the input data is valid.


```{r}

invalid_X <- matrix(c("a","b","c","d"), ncol = 2)

tryCatch(
  k_medoids(invalid_X, k = 2),
  error = function(e) message("Caught expected error: ", e$message)
)

```

<br/>

## 3. Spectral Clustering 


Spectral clustering is especially useful when the true cluster structure is
not linearly separable in the original feature space.

A classical example is **concentric circles**:
the points form two rings around the same center.
A centroid-based method such as k-means often struggles here,
because both rings have the same center, while spectral clustering can use
the graph structure of the data.

### Example: Concentric circles

We first generate a synthetic dataset with two noisy circles.

```{r}
set.seed(123)

n_inner <- 150
n_outer <- 150

theta_inner <- runif(n_inner, 0, 2 * pi)
theta_outer <- runif(n_outer, 0, 2 * pi)

r_inner <- rnorm(n_inner, mean = 1, sd = 0.08)
r_outer <- rnorm(n_outer, mean = 3, sd = 0.12)

inner <- cbind(r_inner * cos(theta_inner), r_inner * sin(theta_inner))
outer <- cbind(r_outer * cos(theta_outer), r_outer * sin(theta_outer))

X <- rbind(inner, outer)
truth <- c(rep(1, n_inner), rep(2, n_outer))

plot(X, col = truth, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "Ground truth",
     cex.main = 1)
```

### Apply k-means

We first apply k-means with two clusters.

```{r}
set.seed(123)
res_km <- k_means(X, K = 2, nstart = 10)

plot(X, col = res_km$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "k-means",
     cex.main = 1)
points(res_km$centers, pch = 8, cex = 1.5)
```

We can already see that k-means tends to split the circles incorrectly,
because it prefers compact spherical clusters around centroids.

### Apply spectral clustering

Now we apply spectral clustering with an RBF affinity graph and the
symmetric normalized Laplacian.

```{r}
res_sc <- spectral_clustering(
  X,
  k = 2,
  affinity = "rbf",
  sigma = 0.6,
  laplacian = "sym",
  normalize_rows = TRUE,
  seed = 123
)

plot(X, col = res_sc$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "Spectral clustering",
     cex.main = 1)
```

### Side-by-side comparison

```{r fig.width=6, fig.height=8}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(X, col = res_km$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "k-means",
     cex.main = 1)
points(res_km$centers, pch = 8, cex = 1.5)

plot(X, col = res_sc$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "Spectral clustering",
     cex.main = 1)

par(mfrow = c(1, 1))
```

The difference is now much clearer:
- `k_means()` typically cuts the data into two halves,
- `spectral_clustering()` can recover the two rings.

### Simple label-invariant accuracy check

Because clustering labels are arbitrary, we compare both possible labelings.

```{r}
cluster_accuracy_2 <- function(pred, truth) {
  acc1 <- mean(pred == truth)
  acc2 <- mean(pred == (3 - truth))
  max(acc1, acc2)
}

acc_km <- cluster_accuracy_2(res_km$clusters, truth)
acc_sc <- cluster_accuracy_2(res_sc$clusters, truth)

acc_km
acc_sc
```

In this example, spectral clustering should achieve a much better result than
k-means, because the graph-based representation captures the ring structure.

### Inspect the spectral embedding

The returned object also contains the low-dimensional embedding used for the
final clustering step.

```{r}
head(res_sc$embedding)
res_sc$eigenvalues
```

This embedding is what makes the nonlinear cluster structure easier to
separate.

<br/>

## 4. Hierarchical Clustering

This shows how to use `hc_hierarchical_clustering()` from the `haufenR` package.
The function builds an agglomerative clustering tree and returns an object of class `"hclust"`.
That means the result can be inspected with the standard `plot()` method and can also be cut into a fixed number of clusters.

### Example data

We create a small data set with three visible groups.

```{r example-data-hc}
set.seed(123)

x <- rbind(
  matrix(rnorm(20, mean = 0, sd = 0.2), ncol = 2),
  matrix(rnorm(20, mean = 3, sd = 0.2), ncol = 2),
  matrix(rnorm(20, mean = 6, sd = 0.2), ncol = 2)
)

plot(x, pch = 19, main = "Example data")
```

### Build the hierarchical tree

Here we use average linkage and directly request a cut into three clusters.

```{r fit-tree}
res <- hc_hierarchical_clustering(x, k = 3, linkage = "average")

res$method
res$dist.method
table(res$clusters)
```

### Plot the dendrogram

Because the returned object has class `"hclust"`, the usual dendrogram plot works directly.

```{r dendrogram}
plot(res, main = "Hierarchical clustering dendrogram", xlab = "Observation")
```

### Inspect the flat clustering

The additional `clusters` element stores the cluster assignment produced by the chosen value of `k`.

```{r inspect-clusters}
head(res$clusters)
length(unique(res$clusters))
```

We# can also color the original points by their assigned cluster.

```{r scatter-plot}
plot(x, col = res$clusters, pch = 19, main = "Cut into 3 clusters")
```

### Compare linkage options

The same data can be clustered with different linkage rules.

```{r linkage-options}
res_single <- hc_hierarchical_clustering(x, linkage = "single")
res_complete <- hc_hierarchical_clustering(x, linkage = "complete")
res_ward <- hc_hierarchical_clustering(x, linkage = "ward")

c(
  single_last_height = tail(res_single$height, 1),
  complete_last_height = tail(res_complete$height, 1),
  ward_last_height = tail(res_ward$height, 1)
)
```

### Remarks

The current implementation supports the Euclidean distance metric and the linkage rules `single`, `complete`, `average`, and `ward`.
Like the other educational implementations in this package, it computes everything directly and is meant to be easy to inspect rather than highly optimized for very large data sets.

<br/>

## 5. DBSCAN/OPTICS

This shows how to use the density-based clustering methods `dbscan()` and `optics()` from the `haufenR` package. The focus is on usage and interpretation of the output. We use a small two-dimensional example data set with two compact groups and two additional points outside these groups.

For a full description of the returned values, see `?dbscan` and `?optics`.
