---
title: "Hierarchical Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load package

```{r setup}
library(haufenR)
```

## Overview

This vignette shows how to use `hc_hierarchical_clustering()` from the `haufenR` package.
The function builds an agglomerative clustering tree and returns an object of class `"hclust"`.
That means the result can be inspected with the standard `plot()` method and can also be cut into a fixed number of clusters.

## Example data

We create a small data set with three visible groups.

```{r example-data}
set.seed(123)

x <- rbind(
  matrix(rnorm(20, mean = 0, sd = 0.2), ncol = 2),
  matrix(rnorm(20, mean = 3, sd = 0.2), ncol = 2),
  matrix(rnorm(20, mean = 6, sd = 0.2), ncol = 2)
)

plot(x, pch = 19, main = "Example data")
```

## Build the hierarchical tree

Here we use average linkage and directly request a cut into three clusters.

```{r fit-tree}
res <- hc_hierarchical_clustering(x, k = 3, linkage = "average")

res$method
res$dist.method
table(res$clusters)
```

## Plot the dendrogram

Because the returned object has class `"hclust"`, the usual dendrogram plot works directly.

```{r dendrogram}
plot(res, main = "Hierarchical clustering dendrogram", xlab = "Observation")
```

## Inspect the flat clustering

The additional `clusters` element stores the cluster assignment produced by the chosen value of `k`.

```{r inspect-clusters}
head(res$clusters)
length(unique(res$clusters))
```

We can also color the original points by their assigned cluster.

```{r scatter-plot}
plot(x, col = res$clusters, pch = 19, main = "Cut into 3 clusters")
```

## Compare linkage options

The same data can be clustered with different linkage rules.

```{r linkage-options}
res_single <- hc_hierarchical_clustering(x, linkage = "single")
res_complete <- hc_hierarchical_clustering(x, linkage = "complete")
res_ward <- hc_hierarchical_clustering(x, linkage = "ward")

c(
  single_last_height = tail(res_single$height, 1),
  complete_last_height = tail(res_complete$height, 1),
  ward_last_height = tail(res_ward$height, 1)
)
```

## Remarks

The current implementation supports the Euclidean distance metric and the linkage rules `single`, `complete`, `average`, and `ward`.
Like the other educational implementations in this package, it computes everything directly and is meant to be easy to inspect rather than highly optimized for very large data sets.
