---
title: "k_medoids_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k_medoids_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(haufenR)
library(cluster)
```

## K-Medoids Clustering

The k-medoids algorithm is a clustering method similar to k-means,
but instead of using the mean of a cluster it chooses actual data points
(medoids) as cluster centers.

This makes the algorithm more robust to outliers compared to k-means.

This vignette demonstrates how to use the k_medoids() function from the
haufenR package and compares it with the pam() implementation from
the cluster package.

## Example 1: Random clusters

We first generate an artificial dataset consisting of three clusters.

```{r}
set.seed(123)

X <- rbind(
  matrix(rnorm(50*2, mean = 0), ncol = 2),
  matrix(rnorm(50*2, mean = 3), ncol = 2),
  matrix(rnorm(50*2, mean = 6), ncol = 2)
)

head(X)
```

Apply our own implementation of the k-medoids algorithm.

```{r}
res <- k_medoids(X, k = 3)

res$medoids
table(res$clusters)

```

## Visualization
```{r}
plot(X, col = res$clusters, pch = 19,
     main = "Own k_medoids implementation", cex.main = 0.8)

points(X[res$medoids, ], pch = 8, cex = 2)

```

The star symbols indicate the selected medoids.

## Example 2: Comparison with PAM

We now compare our implementation with the pam() algorithm from the
cluster package.

```{r}

pam_res <- pam(X, k = 3)

par(mfrow = c(1,2))

plot(X, col = res$clusters, pch = 19,
     main = "Own k_medoids", cex.main = 0.8)
points(X[res$medoids, ], pch = 8, cex = 2)

plot(X, col = pam_res$clustering, pch = 19,
     main = "cluster::pam", cex.main = 0.8)
points(X[pam_res$medoids, ], pch = 8, cex = 2)

par(mfrow = c(1,1))

```

Although the exact cluster assignments may differ due to random
initialization, both algorithms typically produce very similar
cluster structures.

## Example 3: Another dataset

```{r}

set.seed(42)

X2 <- rbind(
  matrix(rnorm(50*2, mean = 0), ncol = 2),
  matrix(rnorm(50*2, mean = 5), ncol = 2)
)

res2 <- k_medoids(X2, k = 2)

plot(X2, col = res2$clusters, pch = 19,
     main = "Two clusters")

points(X2[res2$medoids, ], pch = 8, cex = 2)

```

## Example 4: Single-point dataset

The algorithm also works with minimal datasets.

```{r}

single_point <- matrix(c(1,2), nrow = 1)

res_single <- k_medoids(single_point, k = 1)

res_single$medoids
res_single$clusters

```

## Example 5: Invalid input

Our implementation checks whether the input data is valid.


```{r}

invalid_X <- matrix(c("a","b","c","d"), ncol = 2)

tryCatch(
  k_medoids(invalid_X, k = 2),
  error = function(e) message("Caught expected error: ", e$message)
)

```
