---
title: "k_means_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k_means_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Package "HaufenR"

**Introduction** <br/>
Cluster analysis (German: "Haufenbildung") is a method to discover similarity structures in data. The goal is to group data points so that points within a group are similar to each other and distinct from points in other clusters.


The Package "**haufenR**" implements several clustering methods:<br/>
1. `k_means`: Partitions data into K clusters based on mean values <br/>
2. `k_medoids`: A robust alternative to k-Means, using medoids as cluster centers<br/>
3. `spectral_clustering`: Uses eigenvalues of the similarity matrix to identify clusters<br/>
4. `hc_hierarchical_clustering`: Produces a dendritic structure of clusters<br/>
5. `dbscan`/`optics`: Density-based clustering methods for detecting clusters of arbitrary shape<br/>

```{r setup}
library(haufenR)
```

---

## 1. k-Means
`k-means` partitions numeric data into K clusters by iteratively assigning points to the nearest center and updating the centers until they stabilize. Our `k_means` - function allows multiple random starts to help finding a better solution.

### Example 1: Random Clusters
We generate a random data set
```{r}
set.seed(123)

X <- rbind(matrix(rnorm(50*2, mean = 0), ncol = 2),
           matrix(rnorm(50*2, mean = 3), ncol = 2),
           matrix(rnorm(50*2, mean = 6), ncol = 2))

head(X)
```
Then we apply `k_means` from "haufenR"
```{r}
res <- k_means(X, K = 3, max_iter = 100, nstart = 5)

res$centers

table(res$clusters)
```
Visualize
```{r}
plot(X, col = res$clusters, pch = 19, cex = 0.2, main = "k-means Cluster")
points(res$centers, col = 1:3, pch = 8, cex = 2) # Mark centers
```

### Example 2: Comparison with R's `kmeans`

We use `kmeans` and `k_means` on the same data set and compare
```{r}
set.seed(123)

res_r <- kmeans(X, centers=3) #apply kmeans from R
```

Compare plots
```{r}
par(mfrow = c(1,2)) #Side-by-side plotting

# Visualize own k_mean
plot(X, col = res$clusters, pch = 19, cex = 0.2, main = "Own k_means")
points(res$centers, col = 1:3, pch = 8, cex = 2)

# Visualize R's built in kmean
plot(X, col = res_r$cluster, pch = 19, cex = 0.2, main = "R kmeans")
points(res_r$centers, col = 1:3, pch = 8, cex = 2)
```

Compare centers
```{r}
# Own centers
res$centers
# R's centers
res_r$centers #Same centers (with same seed)

par(mfrow = c(1,1)) #Reset plotting
```

Result: Our own written `k_means` does the same as `kmeans`

### Example 3: Another dataset
```{r}
set.seed(42)
X2 <- rbind(matrix(rnorm(302, mean = 0), ncol = 2),
           matrix(rnorm(502, mean = 5), ncol = 2))
res2 <- k_means(X2, K = 2)
```

Visualize
```{r}
plot(X2, col = res2$clusters, pch=19, cex = 0.2)
points(res2$centers, col = 1:2, pch = 8, cex = 2)
```

### Example 4: Edge case with a single point
```{r}
single_point <- matrix(c(1,2), nrow = 1)
res_single <- k_means(single_point, K = 1)

res_single$centers
res_single$clusters

plot(single_point, col = res_single$clusters, pch = 19, main = "k-means Cluster")
points(res_single$centers, col = 1, pch = 8, cex = 2)
```

### Example 5: Demonstrating multiple random starts improves within-cluster of sum squares
We use `k_means` on a data set with 1 start and with 5 starts
```{r}
set.seed(123)
X3 <- rbind(matrix(rnorm(50*2, mean = 0, sd = 1.5), ncol = 2),
           matrix(rnorm(50*2, mean = 2, sd = 1.5), ncol = 2),
           matrix(rnorm(50*2, mean = 4, sd = 1.5), ncol = 2))
res1 <- k_means(X3, K = 3, nstart = 1)
res5 <- k_means(X3, K = 3, nstart = 5)
```
Compare the tot_withinss
```{r}
res1$tot_withinss
res5$tot_withinss
```
Result: the total amount of distances of the data points to the centers inside each cluster is smaller with a bigger amount of starts, which proves we have have found a way to potentially optimize our code

### Example 6: Invalid input
```{r}
invalid_X <- matrix(c("a", "b", "c", "d"), ncol = 2)
tryCatch(
  k_means(invalid_X, K = 2),
  error = function(e) message("Caught expected error: ", e$message)
)
```
Result: Our Code is able to throw an error if the input is wrong