---
title: "Spectral clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spectral clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load package

```{r setup}
library(haufenR)
```

## Overview

Spectral clustering is especially useful when the true cluster structure is
not linearly separable in the original feature space.

A classical example is **concentric circles**:
the points form two rings around the same center.
A centroid-based method such as k-means often struggles here,
because both rings have the same center, while spectral clustering can use
the graph structure of the data.

## Example: Concentric circles

We first generate a synthetic dataset with two noisy circles.

```{r}
set.seed(123)

n_inner <- 150
n_outer <- 150

theta_inner <- runif(n_inner, 0, 2 * pi)
theta_outer <- runif(n_outer, 0, 2 * pi)

r_inner <- rnorm(n_inner, mean = 1, sd = 0.08)
r_outer <- rnorm(n_outer, mean = 3, sd = 0.12)

inner <- cbind(r_inner * cos(theta_inner), r_inner * sin(theta_inner))
outer <- cbind(r_outer * cos(theta_outer), r_outer * sin(theta_outer))

X <- rbind(inner, outer)
truth <- c(rep(1, n_inner), rep(2, n_outer))

plot(X, col = truth, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "Ground truth",
     cex.main = 1)
```

## Apply k-means

We first apply k-means with two clusters.

```{r}
set.seed(123)
res_km <- k_means(X, K = 2, nstart = 10)

plot(X, col = res_km$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "k-means",
     cex.main = 1)
points(res_km$centers, pch = 8, cex = 1.5)
```

We can already see that k-means tends to split the circles incorrectly,
because it prefers compact spherical clusters around centroids.

## Apply spectral clustering

Now we apply spectral clustering with an RBF affinity graph and the
symmetric normalized Laplacian.

```{r}
res_sc <- spectral_clustering(
  X,
  k = 2,
  affinity = "rbf",
  sigma = 0.6,
  laplacian = "sym",
  normalize_rows = TRUE,
  seed = 123
)

plot(X, col = res_sc$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "Spectral clustering",
     cex.main = 1)
```

## Side-by-side comparison

```r
```{r fig.width=6, fig.height=8}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

plot(X, col = res_km$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "k-means",
     cex.main = 1)
points(res_km$centers, pch = 8, cex = 1.5)

plot(X, col = res_sc$clusters, pch = 19, cex = 0.7, asp = 1,
     xlab = "x1", ylab = "x2",
     main = "Spectral clustering",
     cex.main = 1)

par(mfrow = c(1, 1))
```

The difference is now much clearer:
- `k_means()` typically cuts the data into two halves,
- `spectral_clustering()` can recover the two rings.

## Simple label-invariant accuracy check

Because clustering labels are arbitrary, we compare both possible labelings.

```{r}
cluster_accuracy_2 <- function(pred, truth) {
  acc1 <- mean(pred == truth)
  acc2 <- mean(pred == (3 - truth))
  max(acc1, acc2)
}

acc_km <- cluster_accuracy_2(res_km$clusters, truth)
acc_sc <- cluster_accuracy_2(res_sc$clusters, truth)

acc_km
acc_sc
```

In this example, spectral clustering should achieve a much better result than
k-means, because the graph-based representation captures the ring structure.

## Inspect the spectral embedding

The returned object also contains the low-dimensional embedding used for the
final clustering step.

```{r}
head(res_sc$embedding)
res_sc$eigenvalues
```

This embedding is what makes the nonlinear cluster structure easier to
separate.